GEE models

### 2017-10-03 doc/manuscript.Rmd
UK spelling

Much of the previous experimental work on the role of NEFA in type 2 diabetes 
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid 
(18:1n-9), or alternatively used specific oils such as soybean oil, which is 
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the 
exposure to characterize the metabolic impact of total NEFA. However, fatty
acids comprise multiple molecules with diverse physiological functions, and few 
studies have analysed the effects of a broader spectrum of fatty acids. For instance,
one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta cells [@Kato2008].

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
Participants are contacted annually by telephone. The
current study sample used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterization, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardized across the 2 centres and research
nurses were centrally trained.

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyser and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given their
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see ref @Johnston2016a for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given it's capacity to
handle missed visits. The predictor variables and continuous covariates were
scaled (mean centered and standardized). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis.

@@

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures over time. However, we
believe this is a strength for our specific objective, as the potential impact
that reverse causality, due to the tight integration between fatty acid and
glucose metabolism, may have on the results is lower. This is also an
observational cohort, and there may be some residual confounding we have not
considered or that could not be measured. Nonetheless, potential covariates were
empirically analysed prior to inclusion into the GEE models to best understand
and minimize potential confounding. While there may be potential overadjustment
due to controlling for WC, both QIC and directed acyclic graph outputs
recommended including WC in the GEE model. Even still, we ran sensitivity
analyses and identified WC as the covariate that attenuated the unadjusted
results. This attenuation may be due to the fact that higher NEFA may be 
reflective of greater adipose tissue mass and acting as a mediator between the 
adipose tissue and insulin sensitivity (since adipocytes are active endocrine 
organs).

Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and, while sex and ethnicity did
not influence the results presented from this cohort, our findings may not
entirely be generalizable to other populations. However, given these
limitations, our study also has several strengths, including the longitudinal
design and the rigorous statistical techniques and methods applied, which are
specifically suited to investigating temporal relationships and to handling the
multivariate nature of the data. Lastly, our cohort contains highly detailed and
comprehensive variable measurements at each collection visit, and has both
concentration and mol% data for the fatty acids.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw in general no association of any individual or total NEFA with
hepatic (HOMA2-%S) or whole-body insulin sensitivity (ISI), except for EPA. 
There is some evidence [@Steffen2015a;@Virtanen2014;@Sarbolouki2013] that EPA
may be involved with lowering the risk for diabetes, potentially due to its role
in inflammation [@Calder2009]. However, given we found the association only with
HOMA2-%S and not with ISI, the significant association could be a chance finding,
which will require future studies to confirm. We also identified WC as having a
powerful attenuating effect on the associations between unadjusted and adjusted
models. There are some possible explanations for our null findings for the
insulin sensitivity measures and the influence of WC on the results. First, WC
may be a strong causal link in previously reported associations between NEFA and
insulin sensitivity, given the role adipocytes play in metabolism (e.g. leptin,
adiponectin). Second, there may be differences in physiology between fasting and
postprandial NEFA kinetics that we are not able to investigate but that may
explain our null findings for fasting NEFA and insulin sensitivity. For
instance, some experimental studies using clamp protocols found that fasting
NEFA was a weak predictor of insulin sensitivity compared to postprandial
concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
which may be more metabolically active given postprandial activity. Third, the
null findings seen may be due to the high risk population examined in PROMISE, 
as the majority of subjects had a high WC and BMI. It may be that in this
population, insulin resistance has become well established and NEFA may not
contribute to insulin resistance at this somewhat more advanced stage in the
pathogenesis of diabetes.

>>>>

Much of the previous experimental work on the role of NEFA in type 2 diabetes 
utilised individual fatty acids such as palmitic acid (16:0) or oleic acid 
(18:1n-9), or alternatively used specific oils such as soybean oil, which is 
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the 
exposure to characterise the metabolic impact of total NEFA. However, fatty
acids comprise multiple molecules with diverse physiological functions, and few 
studies have analysed the effects of a broader spectrum of fatty acids. For instance,
one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta cells [@Kato2008].

@@

Participants from London and Toronto, Canada, were recruited into the 
Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
for recruitment into PROMISE required having one or more risk factors for type 2
diabetes mellitus, including obesity, hypertension, family history of diabetes, 
and/or a history of gestational diabetes or birth of a macrosomic infant. 
Participants aged 30 years and older (n=736) attended the baseline visit
between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
Participants are contacted annually by telephone. The
current study sample used data on participants who did not have diabetes at baseline,
who returned for one or more of the follow-up examinations, and who had samples
available for fatty acid measurements (n=477). A diagram of the sample size at
each visit is shown in `r cite_sf('consort')`. At each examination,
participants undergo metabolic characterisation, anthropometric measurements,
and questionnaires on lifestyle and sociodemographics. Research ethics approval
was obtained from Mount Sinai Hospital and the University of Western Ontario,
and all participants provided written informed consent.
Data collection methods were standardised across the 2 centres and research
nurses were centrally trained.

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyser and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorised using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualised under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given their
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see ref @Johnston2016a for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given it's capacity to
handle missed visits. The predictor variables and continuous covariates were
scaled (mean centered and standardised). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis.

@@

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we cannot investigate whether there were 
concomitant changes in NEFA and the metabolic measures over time. However, we
believe this is a strength for our specific objective, as the potential impact
that reverse causality, due to the tight integration between fatty acid and
glucose metabolism, may have on the results is lower. This is also an
observational cohort, and there may be some residual confounding we have not
considered or that could not be measured. Nonetheless, potential covariates were
empirically analysed prior to inclusion into the GEE models to best understand
and minimise potential confounding. While there may be potential overadjustment
due to controlling for WC, both QIC and directed acyclic graph outputs
recommended including WC in the GEE model. Even still, we ran sensitivity
analyses and identified WC as the covariate that attenuated the unadjusted
results. This attenuation may be due to the fact that higher NEFA may be 
reflective of greater adipose tissue mass and acting as a mediator between the 
adipose tissue and insulin sensitivity (since adipocytes are active endocrine 
organs).

Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female and European-ancestry and, while sex and ethnicity did
not influence the results presented from this cohort, our findings may not
entirely be generalisable to other populations. However, given these
limitations, our study also has several strengths, including the longitudinal
design and the rigorous statistical techniques and methods applied, which are
specifically suited to investigating temporal relationships and to handling the
multivariate nature of the data. Lastly, our cohort contains highly detailed and
comprehensive variable measurements at each collection visit, and has both
concentration and mol% data for the fatty acids.

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signalling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
analysis, we saw in general no association of any individual or total NEFA with
hepatic (HOMA2-%S) or whole-body insulin sensitivity (ISI), except for EPA. 
There is some evidence [@Steffen2015a;@Virtanen2014;@Sarbolouki2013] that EPA
may be involved with lowering the risk for diabetes, potentially due to its role
in inflammation [@Calder2009]. However, given we found the association only with
HOMA2-%S and not with ISI, the significant association could be a chance finding,
which will require future studies to confirm. We also identified WC as having a
powerful attenuating effect on the associations between unadjusted and adjusted
models. There are some possible explanations for our null findings for the
insulin sensitivity measures and the influence of WC on the results. First, WC
may be a strong causal link in previously reported associations between NEFA and
insulin sensitivity, given the role adipocytes play in metabolism (e.g. leptin,
adiponectin). Second, there may be differences in physiology between fasting and
postprandial NEFA kinetics that we are not able to investigate but that may
explain our null findings for fasting NEFA and insulin sensitivity. For
instance, some experimental studies using clamp protocols found that fasting
NEFA was a weak predictor of insulin sensitivity compared to postprandial
concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
which may be more metabolically active given postprandial activity. Third, the
null findings seen may be due to the high risk population examined in PROMISE, 
as the majority of subjects had a high WC and BMI. It may be that in this
population, insulin resistance has become well established and NEFA may not
contribute to insulin resistance at this somewhat more advanced stage in the
pathogenesis of diabetes.
